How to build an homogeneous language for heterogeneous platforms

### 2013-05-24 papers/gpce2013/gpce2013.tex
Some changes

\authorinfo{Julien Richard-Foy\and Olivier Barais\and Jean-Marc Jézéquel}
           {IRISA, Université de Rennes 1}
           {\{firstname\}.\{lastname\}@irisa.fr}

@@

\begin{abstract}
This is the text of the abstract.
\end{abstract}

@@

@@

One way consists in hard-coding, in the language's compiler, the transformation to each target platform. Figure
\ref{fat-lang} depicts this process. In order to support a feature related to a specific domain, the whole compiler
pipeline (parser, code generator, \etc) may have to be adapted. This approach gives \emph{fat} languages because
a lot of concepts are defined at the language level. Examples of such languages are Links~\cite{Cooper07_Links},
Opa and Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because each concept is defined in the
compiler, and modifying a compiler requires a high effort. Furthermore, these languages also require to support
common abstraction and composition mechanisms, as general purpose languages do, in order to make it easier to write
real programs. So they need to be at least general purpose languages. It means that for every problem you have to
rewrite a full-featured programming language before addressing the concepts specific to the problem domain. This
approach does not scale.

\begin{figure}
  \centering
  \includegraphics[width=3cm]{fat.pdf}
  \caption{Fat language}
  \label{fat-lang}
\end{figure}

@@

Alternatively, one can define concepts relative to a specific domain as libraries on top of a thin general purpose
language. Figure \ref{thin-lang} depicts this approach. The general purpose language is used as a host language that
does not need to be modified if a new concept is introduced, because concepts are defined as libraries. However,
this approach gives no opportunity to translate a concept efficiently, according to the host platform
characteristics.

\begin{figure}
  \centering
  \includegraphics[width=3cm]{thin.pdf}
  \caption{Thin language}
  \label{thin-lang}
\end{figure}

@@

Figure \ref{dedsl} shows a last approach that tries to get the benefits of the previous two approaches, without
their shortcomings.

\begin{figure}
  \centering
  \includegraphics[width=3cm]{lms.pdf}
  \caption{Deeply embedded languages}
  \label{dedsl}
\end{figure}

@@

We are able to tackle the code sharing issues described in the introduction because of the embdedded nature
of our DSLs: dynamic content of templates is written using embedded DSLs too, so their translation into JavaScript
and Scala is managed by their respective code generators.

@@

@@

Chooze.

\subsection{Several Implementations}

\subsubsection{Vanilla JavaScript}

\subsubsection{jQuery}

\subsubsection{GWT}

\subsubsection{SharpKit}

\subsubsection{Js-Scala}

@@

>>>>

\authorinfo{Julien Richard-Foy\and Olivier Barais\and Jean-Marc Jézéquel}
           {IRISA, Université de Rennes 1}
           {\{first\}.\{last\}@irisa.fr}

@@

\begin{abstract}
Writing code partitioned between heterogeneous client-side and server-side makes the engineering process of Web
applications more tedious. We use a deep embedded DSL to define a common language abstracting over platform
differences for the developers point of view, but generating efficient code leveraging the target platforms
characteristics and giving them control on the amount of code produced on client-side.
\end{abstract}

@@

\begin{figure}
  \centering
  \includegraphics[width=8cm]{langs.pdf}
  \caption{Language engineering processes}
  \label{langs}
\end{figure}

@@

One way consists in hard-coding, in the language's compiler, the transformation to each target platform. Figure
\ref{langs} (left) depicts this process. In order to support a feature related to a specific domain, the whole
compiler pipeline (parser, code generator, \etc) may have to be adapted. This approach gives \emph{fat} languages
because a lot of concepts are defined at the language level. Examples of such languages are
Links~\cite{Cooper07_Links}, Opa and Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because
each concept is defined in the compiler, and modifying a compiler requires a high effort. Furthermore, these
languages also require to support common abstraction and composition mechanisms, as general purpose languages do, in
order to make it easier to write real programs. So they need to be at least general purpose languages. It means that
for every problem you have to rewrite a full-featured programming language before addressing the concepts specific to
the problem domain. This approach does not scale.

@@

Alternatively, one can define concepts relative to a specific domain as libraries on top of a thin general purpose
language. Figure \ref{langs} (middle) depicts this approach. The general purpose language is used as a host language
that does not need to be modified if a new concept is introduced, because concepts are defined as libraries. However,
this approach gives no opportunity to translate a concept efficiently, according to the host platform
characteristics.

@@

Figure \ref{langs} (right) shows a last approach that tries to get the benefits of the previous two approaches,
without their shortcomings.

@@

We are able to tackle the code sharing issues described in the introduction (more details?) because of the embdedded
nature of our DSLs: dynamic content of templates is written using embedded DSLs too, so their translation into
JavaScript and Scala is managed by their respective code generators.

@@

\subsection{Null references}

Code generation consists in traversing the statement nodes produced by the program evaluation according to their
dependencies and to emit the code corresponding to each statement. LMS already sorts the statements graph so DSL
authors just need to say how to emit code for each statement node of their DSL. Listing \ref{option-codegen} shows
such a code generator for the \code{null} reference handling DSL. The \code{emitNode} method handles
\code{OptionIsEmpty} and \code{OptionForeach} nodes. In the case of the \code{OptionIsEmpty} node, it simply
generates an expression testing if the value is \code{null}, in the case of the \code{OptionForeach} node, it wraps
the code block dereferencing the value within a \code{if} checking that the value is not \code{null}.

\begin{figure}
\begin{lstlisting}[caption=Null reference handling DSL code generator,label=option-codegen]
trait JSGenOptionOps extends JSGenEffect {
  val IR: EfffectExp with OptionOpsExp
  import IR._
  override def emitNode(sym: Sym[Any], rhs: Def[Any]) =
    rhs match {
      case OptionIsEmpty(o) =>
        emitValDef(sym, quote(o) + " === null")
      case OptionForeach(o, a, b) =>
        stream.println("if (" + quote(o) + " !== null) {")
        emitValDef(a, quote(o))
        emitBlock(b)
        stream.println("}")
      case _ =>
        super.emitNode(sym, rhs)
    }
}
\end{lstlisting}
\end{figure}

@@

Chooze~\footnote{\href{http://chooze.herokuapp.com}{http://chooze.herokuapp.com}} is an existing complete
application for making polls. It allows users to create a poll, define the choice alternatives, share the poll, vote
and look at the results. It contains some JavaScript code to handle the dynamic behavior of the application:
double-posting prevention, dynamic form update and rich interaction with the document.

The application was initially written using jQuery. We rewrote it using several technologies for the client-side
part: plain JavaScript (without third-party library), js-scala, GWT and SharpKit.

@@

Our goal is to evalute the level of abstraction provided by each solution and their performances. We took the
number of lines of code as a measure of the level of abstraction and we wrote micro-benchmarks to evaluate the
runtime performances of a solution. We also measured the size of the data sent to the client-side. We also measured
the ability to share code between client and server sides.
